Union and intersection contracts are hard, actually

### 2021-06-10 paper.tex
Section 5.1

\subsection{A semantics based on types}

Previous points of this section were about how implementing the expected
semantics for union and intersection has a profound impact on the language. Let
us end the section by investigating a different aspect: does the
semantics correspond to the programmer's intuition?

But which semantics? We deemed the exchange law of Figure~\ref{fig:naive-impl}
as ``clearly wrong'' with respect to overloading. There is nothing wrong per se :
we could choose a semantics for intersections that validate this rule.  However,
the resulting system wouldn't capture overloading, and is arguably not very
interesting nor useful. In~\cite{KeilThiemannUnionIntersection}, the authors
argue for basing the semantics of contracts on union and intersection type
systems, which have a simple and natural definition from which the other
properties follow. This approach makes sense, and has been continued in future
works, and in particular in~\cite{RootCauseOfBlame}.

But even this semantics is no panacea. Figure~\ref{fig:intersection-distribution}
shows an example that intuitively looks like it should succeeed, returning
\nickel{(1, 1)}. However, under the semantics used both
in~\cite{KeilThiemannUnionIntersection} and~\cite{RootCauseOfBlame}, this program
fails with a contract violation.

% Before wrapping up the section, we would like to show that maybe, union and
% intersection contracts are not exactly what we would want them to be.  As stated
% throughout this work, programmers coming from dynamic languages are accustomed
% to perform safety checks (data validation) with great amounts of flexibility and
% the whole power of their language of choice at their disposal.
%
% For instance, a Python programmer may have no problem in checking at runtime,
% that a given variable is an integer if a given flag is not set, as shown in
% Figure \ref{fig:python-assertion}; while a Haskell programmer might close her
% eyes and leave the room instantly.
%
% \begin{figure}[h]
% \begin{lstlisting}[language=python]
% result = 42 if flag else "Disabled"
% ...
% assert (
%   (not flag)
%   or isinstance(result, int)
% )
% \end{lstlisting}
% \caption{Asserting at runtime the type of a variable}
% \label{fig:python-assertion}
% \end{figure}
%
% These expectations are not really met by how unions and intersections
% contracts are supposed to work.
% For example, consider the Nickel code shown on Figure
% \ref{fig:intersection-distrbution}, at first sight, this code should
% complete execution successfully, returning \nickel{(1, 1)},
% however, under both the Keil and Thiemann, as well as the
% Williams, Morris, and Wadler works, this program will fail, with a violation
% of the contract imposed to \nickel{g}.

@@

This discrepancy is inherent to the chosen semantics .  When applying a function
with an intersection contract, only one component of the intersection must be
selected for the function. The actual choice of component may be delayed to
later, but the point is that it must be the same across all usages. In
Figure~\ref{fig:intersection-distribution}, the contract for \nickel{h} can in
consequence be either \nickel{Num -> Num} or \nickel{Bool -> Num}, but not both.
This is a manifestation of the following more general defect: the distributivity
law given in Figure~\ref{fig:distributivity-law} is not valid in the semantics
given in~\cite{KeilThiemannUnionIntersection}, while it is both sound to add and
intuitive.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
A -> B @& A -> C (*@ $\simeq$ @*) A -> (B @& C)
\end{lstlisting}
\caption{Distributivity law}
\label{fig:distributivity-law}
\end{figure}

@@

% LocalWords:  Felleisen JSON TOML YAML templating Jsonnet Dhall MyPy
% LocalWords:  Kubernetes DevOps natively nixpkgs disjunctions
% LocalWords:  TypeScript serializable serialization

>>>>

\subsection{A conductive semantics}

In order to give a precise definition to what values ought to satisfy
union and intersection contracts, \cite{KeilThiemannUnionIntersection}
give a coinductively defined semantics inspired by union and
intersection type systems. The key innovation of
\cite{KeilThiemannUnionIntersection} is recognizing that giving a
semantics to higher-order contracts requires defining not only what
values satisfy a contract, but also what \emph{contexts} satisfy the
contract. This models the situation where context may violate a
contract by calling a function with an inappropriate argument.

Concretely, given a contract $C$, \cite{KeilThiemannUnionIntersection}
introduce the two sets $\llbracket C \rrbracket^+$ and $\llbracket C
\rrbracket^-$ of values and contexts, respectively, satisfying the
contract. They are defined by mutual induction and coinduction.

This semantics has limited support for overloading. Consider the
example in Figure~\ref{fig:intersection-distribution}: it could
evaluate to the pair \nickel{(1, 1)}. But the coinductive semantics
of~\cite{KeilThiemannUnionIntersection} rejects it as a contract
violation. This can be phrased pithily as the fact that the
coinductive semantics doesn't validate the property \nickel{A -> B @&
  A -> C}$\simeq$\nickel{A -> (B @& C)}.

@@

Note that if the last two lines of
Figure~\ref{fig:intersection-distribution} had read \nickel{(g 1 1, g
  1 true)} instead, then the coinductive semantics would accept the
example. It implies that under the coinductive semantics,
common-subexpression elimination (see Section~\ref{sec:optimizations})
is quite perilous.

@@

% LocalWords:  Felleisen JSON TOML YAML templating Jsonnet Dhall MyPy
% LocalWords:  Kubernetes DevOps natively nixpkgs disjunctions
% LocalWords:  TypeScript serializable serialization coinductively
% LocalWords:  coinduction coinductive
